---
title: "20230731_Lab_Teaching_SS"
output: html_document
date: "2023-07-28"
---

# Lab Teaching - Iris

Install and load the necessary packages

```{r}
install.packages(tidyverse)
library(tidyverse)
install.packages(ggpubr)
library(ggpubr)
```

Load Iris dataset

```{r}
data("iris")
view(iris)
```

# Descriptive Statistics - Dataset Characterization 

How many variables do we have? What type of variables do we have?

Categorical Variable: Species

How many iris species were represented in this dataset?

```{r}
iris %>% reframe(unique(Species))
```

How many observations of each iris species were recorded in this dataset?
```{r}
count <- iris %>% group_by(Species) %>% summarise(count = n())
```

Visualizations for Categorical Variables:

Barplot

```{r}
barplot1 <- iris %>% ggplot() + geom_bar(mapping = aes(x = Species, fill = "red"))

barplot2 <- barplot2 <- iris %>% ggplot() + geom_bar(mapping = aes(x = Species, fill = Species)) + scale_fill_manual(values = c("pink", "orange", "purple")) + labs(x = "Species of Iris", y = "Count", title = "Iris Species") + theme_minimal()
```
Pie Chart

```{r}
ggplot(count, aes(x="", y=count, fill=Species)) + geom_bar(stat="identity", width=1) + coord_polar("y", start=0) + theme_void()
```

Qualitative Variables: Petal length and width, sepal length and width

# Distributions of Variables

What are the distributions of these variables?

Histograms

```{r}
petal_length_distribution <- iris %>% ggplot() + geom_histogram(mapping = aes(x = Petal.Length))
petal_width_distribution <- iris %>% ggplot() + geom_histogram(mapping = aes(x = Petal.Width))
sepal_length_distribution <- iris %>% ggplot() + geom_histogram(mapping = aes(x = Sepal.Length))
sepal_width_distribution <- iris %>% ggplot() + geom_histogram(mapping = aes(x = Sepal.Width))
```

QQ plots

```{r}
petal_length_qq <- iris %>% ggplot() + stat_qq(aes(sample = Petal.Length)) + stat_qq_line(aes(sample = Petal.Length))
petal_width_qq <- iris %>% ggplot() + stat_qq(aes(sample = Petal.Width)) + stat_qq_line(aes(sample = Petal.Width))
sepal_length_qq <- iris %>% ggplot() + stat_qq(aes(sample = Sepal.Length)) + stat_qq_line(aes(sample = Sepal.Length))
sepal_width_qq <- iris %>% ggplot() + stat_qq(aes(sample = Sepal.Width)) + stat_qq_line(aes(sample = Sepal.Width))
```

Shapiro-Wilk Normality Test

```{r}
iris %>% pull(Petal.Length) %>% shapiro.test()
iris %>% pull(Petal.Width) %>% shapiro.test()
iris %>% pull(Sepal.Length) %>% shapiro.test()
iris %>% pull(Sepal.Width) %>% shapiro.test()
```

# Calculating Central Tendency and Spread

Normal Distribution:

Mean

```{r}
mean(iris$Sepal.Width)
```

Standard Deviation

```{r}
sd(iris$Sepal.Width)
```

By Species

On average, which species has the widest sepal width? 
```{r}
iris %>% group_by(Species) %>% summarise(mean_sepal_width = mean(Sepal.Width), sd_sepal_width = sd(Sepal.Width)) %>% arrange(desc(mean_sepal_width))
```

Non-Normal Distribution:

Five-Point Summary

```{r}
iris %>% group_by(Species) %>% summarise(min_petal_length = min(Petal.Length), q1_petal_length = quantile(Petal.Length, probs = 0.25), med_petal_length = median(Petal.Length), q3_petal_length = quantile(Petal.Length, probs = 0.75), max_petal_length = max(Petal.Length))

iris %>% summarise(min_petal_length = min(Petal.Length), q1_petal_length = quantile(Petal.Length, probs = 0.25), med_petal_length = median(Petal.Length), q3_petal_length = quantile(Petal.Length, probs = 0.75), max_petal_length = max(Petal.Length))
```

Visualizations:

Boxplots

On average, which species has the longest petal length? 
```{r}
iris %>% ggplot() + geom_boxplot(mapping = aes(x = Petal.Length, y = Species, fill = Species)) + theme_bw()
```

# Statistical Testing

Comparing difference in means between two groups

On average, do virginica irises have longer petal lengths than versicolor and setosa irises?

Pull values as vectors

```{r}
setosa_petal_length <- iris %>% filter(Species == "setosa") %>% pull(Petal.Length) 

versicolor_petal_length <- iris %>% filter(Species == "versicolor") %>% pull(Petal.Length)

virginica_petal_length <- iris %>% filter(Species == "virginica") %>% pull(Petal.Length) 
```

Wilcoxon Rank Sums Test (Non-Parametric)

```{r}
wilcox.test(setosa_petal_length, virginica_petal_length, alternative = "two.sided")

wilcox.test(setosa_petal_length, versicolor_petal_length, alternative = "two.sided")

wilcox.test(virginica_petal_length, versicolor_petal_length, alternative = "two.sided")
```

Welch's Two Sample T Test (Parametric)

```{r}
t.test(setosa_petal_length, virginica_petal_length, alternative = "two.sided")
```

Is there a correlation between petal length and petal width?

Spearman's Correlation (Non-Parametric)

```{r}
cor.test(iris$Petal.Length, iris$Petal.Width, method = "spearman")
```

Pearson's Correlation (Parametric)

```{r}
cor.test(iris$Petal.Length, iris$Petal.Width, method = "pearson")
```

Visualization

Basic Scatter Plot

```{r}
iris %>% ggplot() + geom_point(mapping = aes(x = Petal.Length, y = Petal.Width, color = Species)) + labs(x = "Petal Length", y = "Petal Width") + theme_bw()

iris %>% ggplot() + geom_point(mapping = aes(x = Petal.Length, y = Petal.Width, color = Species, shape = Species)) + labs(x = "Petal Length", y = "Petal Width") + theme_bw() + scale_color_manual(values = c("pink", "orange", "purple"))
```

Spearman's Correlation Plot

```{r}
ggscatter(iris, x = "Petal.Length", y = "Petal.Width", color = "green", add = "reg.line", conf.int = TRUE, cor.coef = TRUE, cor.method = "spearman", xlab = "Petal Length", ylab = "Petal Width", title = "Spearman's Correlation Plot of Iris Petal Length vs. Petal Width") + geom_smooth(method = lm, color = "black")

ggscatter(iris, x = "Petal.Length", y = "Petal.Width", color = "Species", conf.int = TRUE, cor.coef = TRUE, cor.method = "spearman", xlab = "Petal Length", ylab = "Petal Width", title = "Spearman's Correlation Plot of Iris Petal Length vs. Petal Width") + geom_smooth(method = lm, color = "black") + theme(legend.position = "right")

ggscatter(iris, x = "Petal.Length", y = "Petal.Width", color = "Species", add = "reg.line", conf.int = TRUE, cor.coef = TRUE, cor.method = "spearman", xlab = "Petal Length", ylab = "Petal Width", title = "Spearman's Correlation Plot of Iris Petal Length vs. Petal Width") + geom_smooth(method = lm, color = "black") + theme(legend.position = "right")
```
